---
title: "Exploratory Data Analysis Peer Assessment 2"
author: "Alex Konkel"
date: "December 22, 2015"
output: html_document
---

This markdown document will run all the analyses and produce the graphs asked for in the 2nd peer assessment in the John Hopkins Exploratory Data Analysis Coursera class.  The course asks you to copy the relevant code directly into a browser, which seems pretty silly for a data science series that also includes work on reproducible research and sharing data/analyses.  

## Loading the Data
First we need to download the data, unzip it, and load it.
``` {r}
fileName <- c("pm_data.zip")
if(!file.exists(fileName)) {
  download.file("https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip", fileName)
  message("... Unzipping archived source data ...")
  unzip(fileName)
} else {
  message("... Source file already exists ...")
}
pm_data <- readRDS('summarySCC_PM25.rds')
pm_class <- readRDS('Source_Classification_Code.rds')
```
The data seems to load in with reasonable data classes, so we're good to go.

## Question 1: Have Emissions Decreased?
The first part of the assignment is to determine if total emissions from PM2.5 have decreased from 1999 to 2008.  As part of answering the question, we need to make a base graphics plot, saved as a PNG, that shows total emissions for 1999, 2002, 2005, and 2008.  

``` {r}
library(dplyr)
pm_data <- tbl_df(pm_data)
yearly_emissions <- pm_data %>% group_by(year) %>% summarize(total = sum(Emissions))
par(mar=c(4,4,2,0))
with(yearly_emissions,barplot(total,names.arg=c('1999','2002','2005','2008'),xlab='year',ylab='Total PM2.5 Emissions',col='lightblue',main='Emissions by Year Across the US'))
dev.copy(png,file='plot1.png')
dev.off()
```
It looks like emissions have indeed decreased from 1999 to 2008, although there was a lull in the middle where 2002 and 2005 are about the same.  As a more fine-grained check, we can also see if it looks like most counties have had a decline across years.
``` {r}
library(ggplot2)
emissionYearCounty <- pm_data %>% group_by(year,fips) %>% summarize(total = sum(Emissions))
g <- ggplot(emissionYearCounty,aes(as.factor(year),total,group=fips))
g+geom_line()
```

There's one obvious outlier; let's cut that out and focus on the majority of the data
``` {r}
g+geom_line()+coord_cartesian(ylim=c(0,50000))+stat_smooth(method='lm',formula=y~x,se=FALSE)
```

It's still pretty hard to see individual counties, although it seems believable that most show a decrease.  Let's just fit a model and check.
``` {r}
options(scipen=1, digits=2)
library(lme4)
mod1 <- lmer(total ~ year + (year|fips),data=emissionYearCounty)
fixedInt <- summary(mod1)$coef[,1][2]
print(fixedInt)
```
Kind of a rubbish model, but it confirms that the general trend is for emissions to drop.

## Question 2: Have Emissions Dropped in Baltimore?
Now we're going to focus our attention specifically on Baltimore.  Have emissions dropped there?  We'll just redo the earlier code but limited to Baltimore.
``` {r}
BaltByYear <- pm_data %>% group_by(year,fips) %>% summarize(total = sum(Emissions)) %>% filter(fips=='24510')
par(mar=c(4,4,2,0))
barplot(BaltByYear$total,names.arg=c('1999','2002','2005','2008'),xlab='year',ylab='Total PM2.5 Emissions',col='lightblue',main='Emissions by Year in Baltimore')
dev.copy(png,file='plot2.png')
dev.off()
```

It looks like emissions have dropped overall from 1999 to 2008, but there was a spike back up in 2005 before the drop to 2008.

## Question 3: Which Sources Have Changed?
Getting into Baltimore more specifically, let's break it down by pollution source.  Which have declines across the years?  We're supposed to use ggplot for this one.
``` {r}
baltimore <- pm_data %>% filter(fips=='24510')
baltGrouped <- baltimore %>% group_by(year,type) %>% summarize(total=sum(Emissions))
g <- ggplot(baltGrouped,aes(as.factor(year),total,group=type,color=type))
g+geom_line()+labs(x='Year',y='Total PM2.5 Emissions',title='Emissions in Baltimore Over Time by Type')
dev.copy(png,file='plot3.png')
dev.off()
```

It looks like each type has declined overall except for point, which increased but then dropped back to about 1999 levels.

## Question 4: How Has Coal Changed?
Next up: how have emissions from coal sources changed over time, across the entire US?  This is a little tricky because we need to limit to certain sources, but the sources are somewhat heterogeneously named.  For example, in the class file, 'coal' can be labeled 'lignite' in certain levels.  It looks like the EI Sector column is most general and has 'coal' in the entry if coal is involved.  We'll use that with a grep to get the proper SCC code with which to subset the PM data.
``` {r}
coalEntries <- as.character(pm_class$SCC[grep("coal",pm_class$EI.Sector,ignore.case=TRUE)])
coal <- pm_data %>% filter(SCC %in% coalEntries)
```
That should have done our sorting; it looks like not every source has occurred, but that could be the case.  Now let's remake our graph from before, limited to the coal sources.
``` {r}
coalByYear <- coal %>% group_by(year) %>% summarize(total = sum(Emissions))
par(mar=c(4,4,2,0))
barplot(coalByYear$total,names.arg=c('1999','2002','2005','2008'),xlab='year',ylab='Total PM2.5 Emissions',col='lightblue',main='Coal Emissions by Year Across the US')
dev.copy(png,file='plot4.png')
dev.off()
```

It looks like coal emissions have decreased, but almost entirely from 2005 to 2008.

## Question 5: Motor Vehicles in Baltimore
On the topic of specific sources, how have emissions from motor vehicles changed in Baltimore? It looks like the best class variable to sort by is the 'onroad' entry for Data.Category.
``` {r}
carEntries <- as.character(pm_class$SCC[pm_class$Data.Category=='Onroad'])
balCars <- pm_data %>% filter(fips=='24510' & SCC %in% carEntries)
carsByYear <- balCars %>% group_by(year) %>% summarize(total = sum(Emissions))
par(mar=c(4,4,2,0))
barplot(carsByYear$total,names.arg=c('1999','2002','2005','2008'),xlab='year',ylab='Total PM2.5 Emissions',col='lightblue',main='Onroad Emissions by Year in Baltimore')
dev.copy(png,file='plot5.png')
dev.off()
```

It looks like motor vehicle emissions, as indexed by the 'onroad' category, have dropped a fair amount since 1999, although most of it occurred in the 1999-2002 segment.

## Question 6: Baltimore vs. LA
We can see the motor vehicle emissions for Baltimore have decreased, but how do they compare to someplace else?  How about LA?
``` {r}
balCars <- pm_data %>% filter(fips=='24510' & SCC %in% carEntries)
laCars <- pm_data %>% filter(fips=='06037' & SCC %in% carEntries)
balByYear <- balCars %>% group_by(year) %>% summarize(total = sum(Emissions)) %>% mutate(city=rep('Baltimore',4))
laByYear <- laCars %>% group_by(year) %>% summarize(total = sum(Emissions)) %>% mutate(city=rep('LA',4))
carsByYear2 <- bind_rows(balByYear,laByYear)
par(mar=c(4,4,2,0))
g <- ggplot(carsByYear2,aes(year,total,group=city,color=city))
g+geom_line()+labs(x='Year',y='Total PM2.5 Emissions',title='Onroad Emissions in Baltimore vs. LA')
dev.copy(png,file='plot6.png')
dev.off()
```

It looks like LA has much more vehicle missions in general, and also has shown an increase with time while Baltimore's have decreased.  So things aren't necessarily getting better all over.